Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-12 paper.tex

Nonetheless, at least in the single-shared-copy model currently
used by most large-scale variation datasets, Zarr provides
an excellent basis for building a new generation of tools 
that compute \emph{directly} on the data, efficiently
accessing data by variant and by sample.
The vcfzarr specification and scalable vcf2zarr conversion utility
that we have provided here are the starting points of such 
an ecosystem, but much more would be needed to make a viable 
alternative for everyday bioinformatics workflows. However, a few key
pieces of infrastructure would provide a starting point, and would 
be sufficient to greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England.
Firstly, a \texttt{vcfzarrtools} command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) would immediately speed up 
limited region and field specific queries by orders of magnitude,
and would provide compatibility with existing workflows.
Developing such a utility would be a non-trivial but routine
engineering task.

@@

>>>>

Nonetheless, at least in the single-shared-copy model currently
used by most large-scale variation datasets, Zarr provides
an excellent basis for building a new generation of tools 
that compute \emph{directly} on the data, efficiently
accessing data by variant and by sample.
[Mention ``Cloud native data repositories''~\cite{abernathey2021cloud}]
The vcfzarr specification and scalable vcf2zarr conversion utility
that we have provided here are the starting points of such 
an ecosystem, but much more would be needed to make a viable 
alternative for everyday bioinformatics workflows. However, a few key
pieces of infrastructure would provide a starting point, and would 
be sufficient to greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England.
Firstly, a \texttt{vcfzarrtools} command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) would immediately speed up 
limited region and field specific queries by orders of magnitude,
and would provide compatibility with existing workflows.
Developing such a utility would be a non-trivial but routine
engineering task.

@@

Zarr is now widely used.

Multi terabyte numerical weather prediction models~\citep{gowan2022using}
Petabyte scale geoscience datasets in the Pangeo project

OME-Zarr~\cite{rzepka2023toward}
ITS\_LIVE ice velocity dataset~\citep{fahnestock2023mappin} using 
the Zarr.js Javascript interface to visualise the ``massive''
dataset [how big is it?]
